GPU Acceleration of Edge-Based Motion Detection and Machine Learning-Aided Facial Recognition with NVIDIA CUDA

### 2015-12-16 documents/paper/paper.tex
Corrected most grammatical errors in paper

\begin{abstract}
GPUs provide a powerful platform for parallel computations on large data inputs, namely images. In this paper, we explore a GPU-based implementation of a simplified adaptation of existing edge detection algorithms fast enough to operate on frames of a continuous video stream in real-time. We also demonstrate a practical application of edge detection--an edge-based method for motion detection estimation. Additionally, we explore the GPU-CPU speedup of existing OpenCV GPU computation libraries, namely, for facial recognition algorithms. Finally, we demonstrate the speedups as high as 10x we achieve with GPU parallelism, as compared to a reference serial CPU-based implementation.
\end{abstract}

@@

\section{Edge Detection Algorithm}
\label{edge-detection-section}
Edge detection is the process of determining the locations of boundaries that separate objects, or sections of objects, in an image. This edge data simplifies the features of an image to only its basic outlines, which makes it a convenient input for many other problems in computer vision, including motion detection and tracking, object classification, three-dimensional reconstruction, and others. The edge detection algorithm we present is composed of three stages:
\begin{enumerate}
	\item Reduce the amount of high-frequency noise from the image using a two-dimensional low pass filter.
	\item Determine the two-dimensional gradient of the filtered image by applying partial derivatives in both the horizontal and vertical directions.
	\item Classify edges by suppressing surrounding pixels in the gradient direction whose magnitude is lesser, and applying selective thresholding to a limited apron around each selected pixel.
\end{enumerate}
Edges correspond to the points on the image that exhibit a high gradient magnitude. The edge data from our algorithm is then used as input to our motion detection algorithm.

\subsection{Separable Convolution}
The filtering necessary in steps (1) and (2) from our algorithm above involve the convolution of a two-dimensional image and a two-dimensional kernel filter of constant width and height. A na\"{\i}ve implementation of two-dimensional convolution would require a double sum across the kernel and image for each individual pixel:
\begin{equation}
	\label{2d-naive-convolution}
	y[m, n] = x[m, n] * h[m, n] = \sum_{i = -\infty}^{\infty} \sum_{j = -\infty}^{\infty} h[i, j] x[m - i, n - j]
\end{equation}
For an image of size $M \times N$ and kernel of size $k$, a direct implementation would require $O(MNk^2)$ time, which is infeasible for a real-time implementation. Instead, it is possible to exploit the separable property of certain convolution matrices $h$, e.g. that $h$ can be represented as the product of two one-dimensional matrices $h_1$ and $h_2$. This effectively reduces our two-dimensional convolution of Equation \eqref{2d-naive-convolution} to two separate instances of one-dimensional convolution:
\begin{equation}
	\label{separable-convolution-definition}
	y[m, n] = h_1[n] * (h_2[m] * x[m, n])
\end{equation}
which is implemented as
\begin{equation}
	\label{separable-convolution-formula}
	y[m, n] = \sum_{i = -\infty}^{\infty}h_1[i] \sum_{j = -\infty}^{\infty}h_2[j] x[m - i, n - j]
\end{equation}
For the same $M \times N$ input and square kernel of size $k$, a separable implementation reduces the computational complexity to $O(MNk)$. By comparison, an implementation using the FFT would cost $O(MN\log{MN})$. In practice, consider the separable convolution speed gain evident in the following results\footnote{Results were obtained by measuring the CPU execution time of the convolution algorithms implemented in C on an Intel Core i7 4790K @ 4.7 GHz.} for a convolution of a $3 \times 3$ Gaussian filter kernel with images of varying sizes (visualized in Figure \ref{direct-separated-convolution-graph}).
\begin{table}[H]
	\small
	\centering
	\caption{Execution Time of Direct 2D Convolution, $O(MNk^2)$}
	\label{convolution-complexity-comparison}
	\begin{tabular}{llll}
	 Image dimensions & Pixel count ($\times10^6$) & Execution time (s) \\
	 $100 \times 100$ & $0.01$ & $0.000489$ \\
	 $500 \times 500$ & $0.25$ & $0.008784$ \\
	 $1000 \times 1000$ & $1.00$ & $0.034249$ \\
	 $1500 \times 1500$ & $2.25$ & $0.076803$ \\
	 $2000 \times 2000$ & $4.00$ & $0.136222$ \\
	 $2500 \times 2500$ & $6.25$ & $0.212414$ \\
	\end{tabular}
\end{table}
\begin{table}[H]
	\small
	\centering
	\caption{Execution Time of Separated 2D Convolution, $O(MNk)$}
	\label{convolution-complexity-comparison}
	\begin{tabular}{llll}
	 Image dimensions & Pixel count ($\times10^6$) & Execution time (s) \\
	 $100 \times 100$ & $0.01$ & $0.000279$ \\
	 $500 \times 500$ & $0.25$ & $0.004768$ \\
	 $1000 \times 1000$ & $1.00$ & $0.018969$ \\
	 $1500 \times 1500$ & $2.25$ & $0.042410$ \\
	 $2000 \times 2000$ & $4.00$ & $0.075280$ \\
	 $2500 \times 2500$ & $6.25$ & $0.117134$ \\
	\end{tabular}
\end{table}
\begin{figure}
	\centering
	\includegraphics[width=0.45\textwidth]{naive_separable_convolution_execution_time.jpg}
	\caption{Graphical comparison of execution times of direct and separated 2D convolution as a function of the input pixel count}
    \label{direct-separated-convolution-graph}
\end{figure}
As detailed in later sections, we take advantage of the fast computational complexity of separable convolution in our implementation, as all of the filters we apply are separable into one-dimensional matrices.

\subsection{Noise Elimination: Two-Dimensional Gaussian Filtering}
The first task in our edge detection algorithm is denoising the input in order to reduce the amount of high-frequency content in the image by as much as possible without destroying critical information points in the image (e.g. the real edges). We filter out high-frequency noise so that random noise is not mistakenly interpreted as an edge, as edges correspond to points in the image where the gradient has an above-threshold magnitude. \par
For example, consider the $5312 \times 2988$ pixel image (taken at a high ISO to deliberately introduce noise) of Figure \ref{noisy-image} and a plot of its grayscale intensity versus the image's spatial dimensions in Figure \ref{noisy-image-mesh}. It is clear from the mesh that there is much high-frequency noise, which can be removed with a low-pass filter.
\begin{figure}[H]
	\centering
	\includegraphics[width=0.45\textwidth]{noisy_image.jpg}
	\caption{Unfiltered noisy image}
    \label{noisy-image}
\end{figure}
\begin{figure}[h]
	\centering
	\includegraphics[width=0.45\textwidth]{noisy_image_mesh.jpg}
	\caption{Grayscale intensity of unfiltered noisy image at each pixel}
    \label{noisy-image-mesh}
\end{figure}
\begin{figure}[h]
	\centering
	\includegraphics[width=0.45\textwidth]{fft_noisy_image.jpg}
	\caption{FFT magnitude of unfiltered noisy image at each pixel (log scale)}
    \label{noisy-image-fft}
\end{figure}
To low-pass filter our image, we apply a discrete Gaussian filter. Generally, a Gaussian blur kernel of size $2n + 1 \times 2n + 1$ ($\forall n \in \mathbb{Z}^+$, and with parameter $\sigma$) is given by
$$h[i, j] = \frac{1}{2\pi \sigma^2}e^{-\frac{(i - k - 1)^2 + (j - k - 1)^2}{2\sigma^2}}$$
Our implementation of Gaussian filtering uses the constant-sized ($k = 3$), constant-$\sigma$ kernel
\[
h[i, j] = \frac{1}{16}
\begin{bmatrix}
	1 & 2 & 1 \\
	2 & 4 & 2 \\
	1 & 2 & 1
\end{bmatrix}
\]
This kernel is implemented separably as
\begin{equation}
	\label{gaussian-filter}
	h[i, j] = \frac{1}{16}
	\begin{bmatrix}
		1 & 2 & 1
	\end{bmatrix}
	*
	\begin{bmatrix}
		1 \\ 2 \\ 1
	\end{bmatrix}
\end{equation}
\par Applying a Gaussian filter with parameters $k = 5$ and $\sigma = 5$ to the above image significantly denoises the image without sacrificing edge precision, which can be seen from the spatial intensity plot in Figure \ref{filtered-noisy-image-mesh}.
\begin{figure}[H]
	\centering
	\includegraphics[width=0.45\textwidth]{filtered_noisy_image.jpg}
	\caption{Gaussian-filtered noisy image with parameters $k = 3$ and $\sigma = 5$}
    \label{filtered-noisy-image}
\end{figure}
\begin{figure}[h]
	\centering
	\includegraphics[width=0.45\textwidth]{filtered_noisy_image_mesh.jpg}
	\caption{Grayscale intensity of Gaussian-filtered noisy image at each pixel}
    \label{filtered-noisy-image-mesh}
\end{figure}
\begin{figure}[h]
	\centering
	\includegraphics[width=0.45\textwidth]{fft_filtered_noisy_image.jpg}
	\caption{FFT magnitude of Gaussian-filtered noisy image at each pixel (log scale)}
    \label{filtered-noisy-image-fft}
\end{figure}

\subsection{Gradient Computation: The Sobel Operator}
Edges in the image correspond to pixel locations at which there is a rapid change in intensity with respect to the image's spatial dimensions. Thus, edge pixels are defined as those whose gradient magnitude $||\boldsymbol{G}||$ is maximized along the gradient direction $\theta_G$.
\begin{equation}
	\label{gradient-magnitude}
	||\boldsymbol{G}_{i,j}|| = \sqrt{G_x^2 + G_y^2}
\end{equation}
\begin{equation}
	\label{gradient-angle}
	\theta_G = \arctan{\frac{G_y}{G_x}}
\end{equation}
where $G_x$ and $G_y$ are the values of the partial derivatives along the $x$ and $y$ directions, respectively, at the pixel located at $(i, j)$ of the image $A$.
$$G_x = \frac{\partial \boldsymbol{A}}{\partial x}\Bigr|_{\substack{(x, y) = (i, j)}}$$
$$G_y = \frac{\partial \boldsymbol{A}}{\partial y}\Bigr|_{\substack{(x, y) = (i, j)}}$$
\par A variety of different discrete differentiation operators exist to approximate the gradient of the image. The operator we choose in our implementation is the Sobel operator.
\[
\boldsymbol{G_x} =
\begin{bmatrix}
	-1 & 0 & 1 \\
	-2 & 0 & 2 \\
	-1 & 0 & 1
\end{bmatrix} * \boldsymbol{A}
\]
\[
\boldsymbol{G_y} =
\begin{bmatrix}
	-1 & -2 & -1 \\
	0 & 0 & 0 \\
	1 & 2 & 1
\end{bmatrix} * \boldsymbol{A}
\]
Like our Gaussian kernel, the Sobel operator (in both the $x$ and $y$ directions) is separable:
\begin{equation}
	\label{horizontal-sobel-filter}
	\boldsymbol{G_x} =
	\begin{bmatrix}
		1 \\
		2 \\
		1
	\end{bmatrix} * 
	(\begin{bmatrix}
		1 & 0 & -1
	\end{bmatrix} * \boldsymbol{A})
\end{equation}
\begin{equation}
	\label{vertical-sobel-filter}
	\boldsymbol{G_y} =
	\begin{bmatrix}
		1 \\
		0 \\
		-1
	\end{bmatrix} * 
	(\begin{bmatrix}
		1 & 2 & 1
	\end{bmatrix} * \boldsymbol{A})
\end{equation}
\par Thus, we perform two separable convolutions with the separated Sobel filter on the Gaussian-filtered image to obtain gradients in the horizontal and vertical directions, and determine the magnitude and angle matrices $\boldsymbol{G}$ and $\boldsymbol{\theta}$ from equations \eqref{gradient-magnitude} and \eqref{gradient-angle}.
\begin{figure}[h]
	\centering
	\includegraphics[width=0.45\textwidth]{pcb.jpg}
	\caption{Original image input to Sobel filter}
    \label{pcb-original}
\end{figure}
\begin{figure}[h]
	\centering
	\includegraphics[width=0.45\textwidth]{sobel_pcb.jpg}
	\caption{Magnitude of the gradient $||\boldsymbol{G}||$, as determined by applying a Sobel filter in both the $x$ and $y$ directions, at each pixel of the Gaussian-filtered original image with parameters $k = 5$ and $\sigma = 5$}
    \label{sobel-pcb}
\end{figure}
\begin{figure}[h]
	\centering
	\includegraphics[width=0.45\textwidth]{sobel_pcb_mesh.jpg}
	\caption{Gradient of Figure \ref{sobel-pcb} visualized as a three-dimensional mesh}
    \label{sobel-pcb-mesh}
\end{figure}

\subsection{Non-Maximum Suppression and Selective Thresholding}
Given the gradient magnitude and direction of the Gaussian-filtered image, the final step of the edge detection algorithm is to determine which pixels should be selected as edge pixels, and which should be rejected. We define two procedures, \textit{non-maximum suppression} and \textit{selective thresholding}, to gain a reasonably accurate edge map from the gradient.
\par Non-maximum suppression selects and rejects edge pixels according to the following criteria:
\begin{enumerate}
	\item Pixels whose gradient magnitude is below a user-defined low threshold, $t_l$, is immediately rejected
	\item If the gradient magnitude of a pixel is greater than that of the pixels in either direction of the gradient angle of that pixel, then the pixel is accepted if its gradient magnitude is greater than a user-defined high threshold, $t_h$
\end{enumerate}
More precisely, a pixel at $(i, j)$ with gradient magnitude $G_{i, j} > t_l$ and angle $\theta$ is accepted if any of the following hold true:
\begin{itemize}
	\item $(-\frac{\pi}{6} < \theta < \frac{\pi}{6} \vee -\pi < \theta < -\frac{5 \pi}{6} \vee \frac{5\pi}{6} < \theta < \pi) \wedge G_{i, j} > G_{i, j \pm 1}$
	\item $(\frac{\pi}{6} < \theta < \frac{\pi}{3} \vee -\frac{5 \pi}{6} < \theta < -\frac{2 \pi}{3}) \wedge G_{i, j} > G_{i \pm 1, \pm j}$
	\item $(\frac{\pi}{3} < \theta < \frac{2 \pi}{3} \vee -\frac{2 \pi}{3} < \theta < -\frac{\pi}{3}) \wedge G_{i, j} > G_{i \pm 1, j}$
	\item $(\frac{2 \pi}{3} < \theta < \frac{5 \pi}{6} \vee -\frac{\pi}{3} < \theta < -\frac{\pi}{6}) \wedge G_{i, j} > G_{i \pm 1, \mp j}$
\end{itemize}
The selective thresholding technique extends on the procedure of non-maximum suppression. For any pixel that passes criteria (2) (maximized, relative to the neighboring pixels, along the gradient direction), consider an $\alpha$ x $\alpha$ box around the pixel. If any pixel within the box exceeds the low threshold $t_l$, then the pixel is accepted. This technique attempts to catch pixels that might ``connect" ridge pixels identified by criteria (2) but fail to satisfy the criterion themselves, in order to create more continuous edge lines.
\begin{figure}[]
	\centering
	\includegraphics[width=0.45\textwidth]{pcb_edges.jpg}
	\caption{Output of edge detection algorithm after non-maximum suppression and selective thresholding on the image of Figure \ref{pcb-original}, with parameters $t_l = 70$ and $t_h = 80$}
    \label{pcb-edges}
\end{figure}
\begin{algorithm}[h]
	\smaller
	\label{non-maximum-suppression-selective-thresholding-algorithm}
	\caption{Non-maximum suppression and selective thresholding: pseudocode for a serial, CPU implementation}

@@

\subsection{GPU Parallelization and CUDA Implementation}
We begin with a parallelized implementation of separable convolution, since the same convolution operation is applied for both the initial Gaussian low pass filter and the edge detecting Sobel filter. Our parallel implementation involves first independently computing the convolution with the horizontal filter ($h_2$ in Equation \eqref{separable-convolution-definition}), then convolving that result with the vertical filter $h_1$. In computing the output $y[m, n]$ at every index $(i, j)$, we launch a kernel with 16 threads per block, where the total number of blocks along one dimension is equal to the size of that dimension divided by 16. For our 5312 x 2988 test images, this corresponds to 332 blocks in the horizontal direction and 186 blocks in the vertical direction, each with 16 computational threads.
\begin{lstlisting}[
	caption=Grid configuration for the separable convolution device kernel,
	label=grid-configuration,
	language=C
]
#define TX 16
#define TY 16

@@

\section{Facial Recognition Algorithm}
\label{facial-recognition-section}
Our facial recognition implementation is based off of the HAAR features-based cascade classifiers method proposed by Viola-Jones in 2001. It's split into two phases: training and detection. The algorithm is as follows:
\begin{enumerate}
	\item Extract HAAR features of a face from a set of positive and negative training images. 
	\item Use adaptive boosting machine learning to train a multi-stage, or cascade, classifier against positive and negative HAAR features. 
	\item Apply a final cascade classifier to a loaded image or real time video frame. 
\end{enumerate}
Due to the resource capacity necessary to train a full facial recognition classifier, we used a pre-trained frontal face cascade classifier and used OpenCV's object detection library, which encapsulates the algorithm above, and whose theory detailed in the following sections.

>>>>

\begin{abstract}
GPUs provide a powerful platform for parallel computations on large data inputs, namely images. In this paper, we explore a GPU-based implementation of a simplified adaptation of existing edge detection algorithms fast enough to operate on frames of a continuous video stream in real-time. We also demonstrate a practical application of edge detection--an edge-based method for motion detection estimation. Additionally, we explore the GPU-CPU speedup of existing OpenCV GPU computation libraries, namely, for facial recognition algorithms. Finally, we demonstrate speedups as high as 10x we achieve with GPU parallelism, as compared to a reference serial CPU-based implementation.
\end{abstract}

@@

\section{Edge Detection Algorithm}
\label{edge-detection-section}
Edge detection is the process of determining the locations of boundaries that separate objects, or sections of objects, in an image. This edge data simplifies the features of an image to only its basic outlines, which makes it a convenient input for many other problems in computer vision, including motion detection and tracking, object classification, three-dimensional reconstruction, and others. The edge detection algorithm we present is composed of three stages:
\begin{enumerate}
	\item Reduce the amount of high-frequency noise from the image using a two-dimensional low pass filter.
	\item Determine the two-dimensional gradient of the filtered image by applying partial derivatives in both the horizontal and vertical directions.
	\item Classify edges by accepting pixels of a minimum threshold gradient magnitude and suppressing its neighboring pixels in the gradient direction with a lesser gradient magnitude, and applying selective thresholding to a limited apron around each selected pixel.
\end{enumerate}
Edges correspond to the points on the image that exhibit a high gradient magnitude. The edge data from our algorithm is then used as input to our motion detection algorithm.

\subsection{Separable Convolution}
The filtering necessary in steps (1) and (2) from our algorithm above involve the convolution of a two-dimensional image and a two-dimensional kernel filter of constant width and height. A na\"{\i}ve implementation of two-dimensional convolution would require a double sum across the convolution kernel and input image for each individual pixel in the output image:
\begin{equation}
	\label{2d-naive-convolution}
	y[m, n] = x[m, n] * h[m, n] = \sum_{i = -\infty}^{\infty} \sum_{j = -\infty}^{\infty} h[i, j] x[m - i, n - j]
\end{equation}
For an image of size $M \times N$ and kernel of size $k$, a direct implementation would require $O(MNk^2)$ time, which is infeasible for a real-time implementation. Instead, it is possible to exploit the separable property of certain convolution matrices $h$, e.g. that $h$ can be represented as the product of two one-dimensional matrices $h_1$ and $h_2$. This effectively reduces our two-dimensional convolution of Equation \eqref{2d-naive-convolution} to two separate instances of one-dimensional convolution:
\begin{equation}
	\label{separable-convolution-definition}
	y[m, n] = h_1[n] * (h_2[m] * x[m, n])
\end{equation}
which is implemented as
\begin{equation}
	\label{separable-convolution-formula}
	y[m, n] = \sum_{i = -\infty}^{\infty}h_1[i] \sum_{j = -\infty}^{\infty}h_2[j] x[m - i, n - j]
\end{equation}
For the same $M \times N$ input and square kernel of size $k$, a separable implementation reduces the computational complexity to $O(MNk)$. By comparison, an implementation using the FFT would cost $O(MN\log{MN})$. In practice, consider the separable convolution speed gain evident in the following results\footnote{Results were obtained by measuring the CPU execution time of the convolution algorithms implemented in C on an Intel Core i7 4790K @ 4.7 GHz.} for a convolution of a $3 \times 3$ Gaussian filter kernel with images of varying sizes (visualized in Figure \ref{direct-separated-convolution-graph}).
\begin{table}[H]
	\small
	\centering
	\caption{Execution Time of Direct 2D Convolution, $O(MNk^2)$}
	\label{convolution-complexity-comparison}
	\begin{tabular}{llll}
	 Image dimensions & Pixel count ($\times10^6$) & Execution time (s) \\
	 $100 \times 100$ & $0.01$ & $0.000489$ \\
	 $500 \times 500$ & $0.25$ & $0.008784$ \\
	 $1000 \times 1000$ & $1.00$ & $0.034249$ \\
	 $1500 \times 1500$ & $2.25$ & $0.076803$ \\
	 $2000 \times 2000$ & $4.00$ & $0.136222$ \\
	 $2500 \times 2500$ & $6.25$ & $0.212414$ \\
	\end{tabular}
\end{table}
\begin{table}[H]
	\small
	\centering
	\caption{Execution Time of Separated 2D Convolution, $O(MNk)$}
	\label{convolution-complexity-comparison}
	\begin{tabular}{llll}
	 Image dimensions & Pixel count ($\times10^6$) & Execution time (s) \\
	 $100 \times 100$ & $0.01$ & $0.000279$ \\
	 $500 \times 500$ & $0.25$ & $0.004768$ \\
	 $1000 \times 1000$ & $1.00$ & $0.018969$ \\
	 $1500 \times 1500$ & $2.25$ & $0.042410$ \\
	 $2000 \times 2000$ & $4.00$ & $0.075280$ \\
	 $2500 \times 2500$ & $6.25$ & $0.117134$ \\
	\end{tabular}
\end{table}
\begin{figure}
	\centering
	\includegraphics[width=0.45\textwidth]{naive_separable_convolution_execution_time.jpg}
	\caption{Graphical comparison of execution times of direct and separated 2D convolution as a function of the input pixel count}
    \label{direct-separated-convolution-graph}
\end{figure}
As detailed in later sections, we take advantage of the fast computational complexity of separable convolution in our implementation, as all of the filters we apply are separable into one-dimensional matrices.

\subsection{Noise Elimination: Two-Dimensional Gaussian Filtering}
The first task in our edge detection algorithm is denoising the input in order to reduce the amount of high-frequency content in the image by as much as possible without destroying critical information points in the image (e.g. the real edges). We filter out high-frequency noise so that random noise is not mistakenly interpreted as an edge, as edges correspond to points in the image where the gradient has an above-threshold magnitude. \par
For example, consider the $5312 \times 2988$ pixel image (taken at a high ISO to deliberately introduce noise) of Figure \ref{noisy-image} and a plot of its grayscale intensity versus the image's spatial dimensions in Figure \ref{noisy-image-mesh}. It is clear from the mesh that there exists much high-frequency noise, which can be removed with a low-pass filter.
\begin{figure}[H]
	\centering
	\includegraphics[width=0.45\textwidth]{noisy_image.jpg}
	\caption{Unfiltered noisy image}
    \label{noisy-image}
\end{figure}
\begin{figure}[h]
	\centering
	\includegraphics[width=0.45\textwidth]{noisy_image_mesh.jpg}
	\caption{Grayscale intensity of the unfiltered noisy image of Figure \ref{noisy-image} at each pixel, visualized as a three-dimensional mesh}
    \label{noisy-image-mesh}
\end{figure}
\begin{figure}[h]
	\centering
	\includegraphics[width=0.45\textwidth]{fft_noisy_image.jpg}
	\caption{FFT magnitude of the unfiltered noisy image of Figure \ref{noisy-image} at each pixel (log scale)}
    \label{noisy-image-fft}
\end{figure}
To low-pass filter our image, we apply a discrete Gaussian filter. Generally, a Gaussian blur kernel of size $2n + 1 \times 2n + 1$ ($\forall n \in \mathbb{Z}^+$, and with parameter $\sigma$) is given by
$$h[i, j] = \frac{1}{2\pi \sigma^2}e^{-\frac{(i - k - 1)^2 + (j - k - 1)^2}{2\sigma^2}}$$
Our implementation of Gaussian filtering uses the constant-sized ($k = 3$), constant-$\sigma$ kernel
\[
h[i, j] = \frac{1}{16}
\begin{bmatrix}
	1 & 2 & 1 \\
	2 & 4 & 2 \\
	1 & 2 & 1
\end{bmatrix}
\]
This kernel is implemented separably as
\begin{equation}
	\label{gaussian-filter}
	h[i, j] = \frac{1}{16}
	\begin{bmatrix}
		1 & 2 & 1
	\end{bmatrix}
	*
	\begin{bmatrix}
		1 \\ 2 \\ 1
	\end{bmatrix}
\end{equation}
\par Applying a Gaussian filter with parameters $k = 5$ and $\sigma = 5$ to the noisy image of Figure \ref{noisy-image} significantly denoises the image without sacrificing edge precision, which can be seen from the spatial intensity plot in Figure \ref{filtered-noisy-image-mesh}.
\begin{figure}[H]
	\centering
	\includegraphics[width=0.45\textwidth]{filtered_noisy_image.jpg}
	\caption{Gaussian-filtered noisy image with parameters $k = 3$ and $\sigma = 5$}
    \label{filtered-noisy-image}
\end{figure}
\begin{figure}[h]
	\centering
	\includegraphics[width=0.45\textwidth]{filtered_noisy_image_mesh.jpg}
	\caption{Grayscale intensity of the Gaussian-filtered noisy image of Figure \ref{filtered-noisy-image} at each pixel, visualized as a three-dimensional mesh}
    \label{filtered-noisy-image-mesh}
\end{figure}
\begin{figure}[h]
	\centering
	\includegraphics[width=0.45\textwidth]{fft_filtered_noisy_image.jpg}
	\caption{FFT magnitude of the Gaussian-filtered noisy image of Figure \ref{filtered-noisy-image} at each pixel (log scale)}
    \label{filtered-noisy-image-fft}
\end{figure}

\subsection{Gradient Computation: The Sobel Operator}
Edges in the image correspond to pixel locations at which there is a rapid change in intensity with respect to the image's spatial dimensions. Thus, edge pixels are defined as those whose gradient magnitude $||\boldsymbol{G}||$ is maximized along the gradient direction $\theta_G$.
\begin{equation}
	\label{gradient-magnitude}
	||\boldsymbol{G}_{i,j}|| = \sqrt{G_x^2 + G_y^2}
\end{equation}
\begin{equation}
	\label{gradient-angle}
	\theta_G = \arctan{\frac{G_y}{G_x}}
\end{equation}
where $G_x$ and $G_y$ are the values of the partial derivatives along the $x$ and $y$ directions, respectively, at the pixel located at $(i, j)$ of the image $\boldsymbol{A}$.
$$G_x = \frac{\partial \boldsymbol{A}}{\partial x}\Bigr|_{\substack{(x, y) = (i, j)}}$$
$$G_y = \frac{\partial \boldsymbol{A}}{\partial y}\Bigr|_{\substack{(x, y) = (i, j)}}$$
\par A variety of different discrete differentiation operators exist to approximate the gradient of the image. The operator we choose in our implementation is the Sobel operator.
\[
\boldsymbol{G_x} =
\begin{bmatrix}
	-1 & 0 & 1 \\
	-2 & 0 & 2 \\
	-1 & 0 & 1
\end{bmatrix} * \boldsymbol{A}
\]
\[
\boldsymbol{G_y} =
\begin{bmatrix}
	-1 & -2 & -1 \\
	0 & 0 & 0 \\
	1 & 2 & 1
\end{bmatrix} * \boldsymbol{A}
\]
Like our Gaussian kernel, the Sobel operator (in both the $x$ and $y$ directions) is separable:
\begin{equation}
	\label{horizontal-sobel-filter}
	\boldsymbol{G_x} =
	\begin{bmatrix}
		1 \\
		2 \\
		1
	\end{bmatrix} * 
	(\begin{bmatrix}
		1 & 0 & -1
	\end{bmatrix} * \boldsymbol{A})
\end{equation}
\begin{equation}
	\label{vertical-sobel-filter}
	\boldsymbol{G_y} =
	\begin{bmatrix}
		1 \\
		0 \\
		-1
	\end{bmatrix} * 
	(\begin{bmatrix}
		1 & 2 & 1
	\end{bmatrix} * \boldsymbol{A})
\end{equation}
\par Thus, we perform two separable convolutions with the separated Sobel filter on the Gaussian-filtered image to obtain gradients in the horizontal and vertical directions, and determine the magnitude and angle matrices $\boldsymbol{G}$ and $\boldsymbol{\theta}$ from equations \eqref{gradient-magnitude} and \eqref{gradient-angle}.
\begin{figure}[h]
	\centering
	\includegraphics[width=0.45\textwidth]{pcb.jpg}
	\caption{Original image input to Sobel filter}
    \label{pcb-original}
\end{figure}
\begin{figure}[h]
	\centering
	\includegraphics[width=0.45\textwidth]{sobel_pcb.jpg}
	\caption{Magnitude of the gradient $||\boldsymbol{G}||$ as determined by applying a Sobel filter in both the $x$ and $y$ directions to the Gaussian-filtered original image of Figure \ref{pcb-original} with parameters $k = 5$ and $\sigma = 5$}
    \label{sobel-pcb}
\end{figure}
\begin{figure}[h]
	\centering
	\includegraphics[width=0.45\textwidth]{sobel_pcb_mesh.jpg}
	\caption{Gradient magnitude of Figure \ref{sobel-pcb} visualized as a three-dimensional mesh}
    \label{sobel-pcb-mesh}
\end{figure}

\subsection{Non-Maximum Suppression and Selective Thresholding}
Given the gradient magnitude and direction of the Gaussian-filtered image, the final step of the edge detection algorithm is to determine which pixels should be selected as edge pixels, and which should be rejected. We define two procedures, \textit{non-maximum suppression} and \textit{selective thresholding}, to gain a reasonably accurate edge map from the gradient.
\par Non-maximum suppression selects and rejects edge pixels according to the following criteria:
\begin{enumerate}
	\item Pixels whose gradient magnitude is below a user-defined low threshold, $t_l$, is immediately rejected
	\item If the gradient magnitude of a pixel is greater than that of the pixels in either direction of the gradient angle of that pixel, then the pixel is accepted if its gradient magnitude is greater than a user-defined high threshold, $t_h$
\end{enumerate}
More precisely, a pixel at $(i, j)$ with gradient magnitude $G_{i, j} > t_l$ and angle $\theta$ is accepted if any of the following hold true:
\begin{itemize}
	\item $(-\frac{\pi}{6} < \theta < \frac{\pi}{6} \vee -\pi < \theta < -\frac{5 \pi}{6} \vee \frac{5\pi}{6} < \theta < \pi) \wedge G_{i, j} > G_{i, j \pm 1}$
	\item $(\frac{\pi}{6} < \theta < \frac{\pi}{3} \vee -\frac{5 \pi}{6} < \theta < -\frac{2 \pi}{3}) \wedge G_{i, j} > G_{i \pm 1, \pm j}$
	\item $(\frac{\pi}{3} < \theta < \frac{2 \pi}{3} \vee -\frac{2 \pi}{3} < \theta < -\frac{\pi}{3}) \wedge G_{i, j} > G_{i \pm 1, j}$
	\item $(\frac{2 \pi}{3} < \theta < \frac{5 \pi}{6} \vee -\frac{\pi}{3} < \theta < -\frac{\pi}{6}) \wedge G_{i, j} > G_{i \pm 1, \mp j}$
\end{itemize}
\par The selective thresholding technique extends on the procedure of non-maximum suppression. For any pixel that passes criteria (2) (maximized, relative to the neighboring pixels, along the gradient direction), consider an $\alpha \times \alpha$ box around the pixel. If any pixel within the box exceeds the low threshold $t_l$, then the pixel is accepted. This technique attempts to catch pixels that might ``connect" ridge pixels identified by criteria (2) but fail to satisfy the criterion themselves, in order to create more continuous edge lines.
\begin{figure}[]
	\centering
	\includegraphics[width=0.45\textwidth]{pcb_edges.jpg}
	\caption{Output of edge detection algorithm after non-maximum suppression and selective thresholding on the image of Figure \ref{pcb-original}, with parameters $t_l = 70$ and $t_h = 80$}
    \label{pcb-edges}
\end{figure}
\begin{algorithm}[h]
	\smaller
	\label{non-maximum-suppression-selective-thresholding-algorithm}
	\caption{Non-maximum suppression and selective thresholding: pseudocode for a serial, CPU implementation}

@@

\subsection{GPU Parallelization and CUDA Implementation}
We begin with a parallelized implementation of separable convolution, since the same convolution operation is applied for both the initial Gaussian low pass filter and the edge detecting Sobel filter. Our parallel implementation involves first independently computing the convolution with the horizontal filter ($h_2$ in Equation \eqref{separable-convolution-definition}), then convolving that result with the vertical filter $h_1$. In computing the output $y[m, n]$ at every index $(i, j)$, we launch a kernel with 16 threads per block, where the total number of blocks along one dimension is equal to the size of that dimension divided by 16. For our 5312 $\times$ 2988 test images, this corresponds to 332 blocks in the horizontal direction and 186 blocks in the vertical direction, each with 16 computational threads.
\begin{lstlisting}[
	caption=Grid configuration for the separable convolution device kernel,
	label=grid-configuration,
	language=C
]
#define TX 16
#define TY 16

@@

\section{Facial Recognition Algorithm}
\label{facial-recognition-section}
Our facial recognition implementation is based off the HAAR feature-based cascade classifiers method proposed by Viola-Jones in 2001. It's split into two phases: training and detection. The algorithm is as follows:
\begin{enumerate}
	\item Extract HAAR features of a face from a set of positive and negative training images. 
	\item Use adaptive boosting machine learning to train a multi-stage, or cascade, classifier against positive and negative HAAR features. 
	\item Apply a final cascade classifier to a loaded image or real time video frame. 
\end{enumerate}
Due to the resource capacity necessary to train a full facial recognition classifier, we used a pre-trained frontal face cascade classifier and used OpenCV's object detection library, which encapsulates the algorithm above, and whose theory detailed in the following sections.

\subsection{HAAR Feature Extraction}
In mathematics analysis, Haar wavelets are a sequence of rescaled, square-like functions. In image processing, Haar features are so named for their resemblance to Haar wavelets. Haar features are regularities identified in all human faces, by comparing light and dark gradients across the face. These are split into edge (two-rectangle) features, line (three-rectangle) features, and diagonal (four-rectangle) features.
\begin{figure}[h]
	\centering
	\includegraphics[width=0.45\textwidth]{haar_features.png}
	\caption{Example application of HAAR features on a face}
    \label{haar-features-face-example}
\end{figure}
\par Each feature is obtained by comparing the pixel intensity in the white boxes against the pixel intensity in the dark boxes for a wide sample of sub-windows across the entire image. A 24 $\times$ 24 pixel image window can produce over 160,000 features that need to be classified--in some sense, “oversampling” the face.
\par The rectangle features can be computed by splitting the image into sub-windows called integral images. An integral image at a point $i'(x, y)$ is the sum of all pixels to the top and to the left of that point.
\[
i'(x, y) = \sum_{m \leq x, n \leq y}{i(m, n)}
\]
Since an integral image includes all previously-computed integral images contained in its window, in the image of Figure \ref{integral-image-face}, the region A can be computed as
\[
A = i'(4) + i'(1) - i'(2) - i'(3)
\]
Any such rectangular sum can be computed using four array references.
\begin{figure}[h]
	\centering
	\includegraphics[width=0.45\textwidth]{integral_image.png}
	\caption{Integral image on a face}
    \label{integral-image-face}
\end{figure}
